Author Guidelines for DAGM-OAGM Submission

### 2013-09-27 itm/paper.tex
k-means not capitalizes, minor fixes

\section{Related Work}
The most commonly used clustering algorithm is the $k$-Means algorithm, originally
investigated by \citet{macqueen1967some} and most commonly implemented using
Lloyd's algorithm \citep{macqueen1967some,lloyd1982least}. While
$k$-Means often works well in practice, one of its main drawbacks is the
restriction in cluster shape. 
%
Clusters are given by the Voronoi tessellation of the cluster means and
therefore are always convex.  Another drawback is the non-determinism of the
procedure, caused by the dependence on random initialization.

Another widely used method is spectral clustering~\citep{shi2000normalized,ng2002spectral}, 
which solves a graph partitioning problem on a similarity
graph constructed from the data. While spectral clustering is much more
flexible than $k$-Means it is quite sensitive to the particular choice of graph
construction and similarity measure. It is also computationally expensive to
compute, because clustering $n$ points requires computing the eigenvalues and
-vectors of an $n{\times}n$ matrix.

@@

Later, \citet{banerjee2005clustering} introduced the concept of Bregman
Information, generalizing mutual information of distributions, and showed how
this leads to a natural formulation of several clustering algorithms.
%
\citet{barber2006kernelized} constructed a soft clustering by using a parametric model of $p(Y\given X)$.
%
The framework of mutual information based clustering was extended to
non-parametric entropy estimates by \citet{faivishevsky2010nonparametric}.
They use a nearest neighbor based estimator of the mutual information, called
MeanNN, that takes into account all possible neighborhoods, therefore combining
global and local influences. The approximate mutual information is maximized
using local search over labels.
%discuss this later in detail
%While optimizing MeanNN allows for non-convex
%clusters, in practice the result is strongly biased towards compact, convex
%regions. This can be explained by a close connections of MeanNN to $k$-Means,
%see Section~\ref{discussion} for an in-depth discussion.

Clustering algorithms based on minimum spanning trees have been studied early
on in the statistics community, due to their efficiency.  One of the earliest
methods is single-link agglomerative clustering~\citep{gower1969minimum}.
Single-link agglomerative clustering can be understood as a minimum spanning
tree-based approach in which the largest edge is removed until the desired
number of components is reached.  \citet{zahn1971graph} refined this criterion
by cutting edges that are longer than other edges in the vicinity. This
approach requires tuning several constants by hand. More recently,
\citet{grygorash2006minimum} proposed a hierarchical MST-based clustering
approach that iteratively cuts edges, merges points in the resulting
components, and rebuilds the spanning tree.
We limit our discussion to the most widely used algorithm from~\citep{gower1969minimum}.
%
\section[Clustering Using Non-Parametric Entropy Estimates]{Clustering Using\linebreak Non-Parametric Entropy Estimates}
In general, the goal of clustering can be formulated as follows: 
given a finite collection of samples $\mathbf{x} = (x_1, \dotsc, x_n)$, we want to 
assign cluster-memberships $\mathbf{y} = (y_1, \dotsc, y_n ), y_i \in \{1, \dotsc k\}$ to these samples.
%
We adopt the viewpoint of information theoretic clustering of \citet{gokcay2002information}, 
where the $x_i$ are considered i.i.d{.} samples from a distribution $p(X)$, and the 
$y_i$ are found such that the mutual information $I(X, Y)$ between the distribution $p(X)$ and the
assigned labels $p(Y)$ is maximized.
We can rewrite this objective as 
\begin{align}
         I(X, Y)
        &= D_{\text{KL}}\bigl(p(X, Y) \parallel p(X) p(Y)\bigr) = H(X) - \sum_{y=1}^k p(Y{=}y) H(X \given Y {=} y)
\end{align}
where
\begin{itemize}
    \item $\displaystyle D_\text{KL}\bigl(p(X) \parallel q(X)\bigr)\, = \int_\mathcal{X} p(X) \ln\left(\frac{p(X)}{q(X)}\right)\,dX$ is the Kullback-Leibler divergence,
    \item $\displaystyle H(X)\, = \int_\mathcal{X} p(X) \ln\bigl(p(X)\bigr)\,dX$ is the differential entropy, and
    \item $\displaystyle H(X \given Y{=}y)\, = \int_\mathcal{X} p(X \given Y {=}y) \ln\bigl(p(X \given Y {=} y)\bigr)\,dX$ is the conditional differential entropy.
    \end{itemize}
Expressing the mutual information in terms of the entropy is convenient, since
the objective then decomposes over the values of $Y$. 
%
Additionally, $H(X)$ is independent of the distribution of $Y$ and therefore
does not influence the search over $\mathbf{y}$.

@@

We use Prim's algorithm combined with a ball tree data structure for distance
queries to compute the minimum spanning tree of the data.
While this procedure has no strong runtime guarantees, we found this faster
in practice than specialized algorithms for euclidean minimum spanning trees, 
which achieve a better theoretical runtime of $O\bigl(n \log(n) \alpha(n)\bigr)$ \citep{dtb2010}. Here
$\alpha$ is the inverse of the Ackerman function. The dynamic programming solution of
Algorithm~\ref{overall} has a run time of $O(n)$ per removed edge, leading to an overall
run time of $O\bigl(n \log(n) \alpha(n)+ nk\bigr)$. The $O(nk)$ comes from a worst case
scenario, in which each step in the hierarchical clustering procedure only
splits off a constant number of points. In a more realistic setting, we expect
that the individual clusters are much smaller than the original dataset. In
this case, the $O(nk)$ factor would improve to $O\bigl(n\log(k)\bigr)$.

@@

A particularly drastic example is the case of the dataset having less samples
than features. In this case, the data clearly lies even on a linear subspace
of the input space, and the dimensionality of the input space does not
accurately reflect the intrinsic dimensionality of the data.
We use the estimate analyzed in \citet{massoud2007manifold}.
In their method, for each data point $x$, a local estimate $\hat{d}(x)$ of the
dimensionality at $x$ is computed as
\begin{equation}
    \hat{d}(x) = \frac{\ln 2}{\ln\left(r_k(x) / r_{\lfloor k /2 \rfloor}(x)\right)}.
\end{equation}
Here $k$ is a fixed integer and $r_k(x)$ is the distance of $x$ from its $k$th neighbor.
We follow \citet{massoud2007manifold} and set $k =\lceil 2 \ln n\rceil$.
The final estimate $\hat{d}$ is then computed by averaging the estimates over all $x$
\begin{equation}
    \hat{d} = \frac{1}{n}\sum_{x \in X} \min\left(\hat{d}(x),\ d\right).
\end{equation}
We compute the density estimate once, prior to clustering, and then plug the
estimate $\hat{d}$ into \eqref{graphobjective} in place of $d$.
We found this estimate to work robustly and give sensible results for all
datasets we investigated. As we already used a ball tree data structure to
build the minimum spanning trees, we can reuse this structure to compute $r_k$.
Consequently, estimating the dimensionality resulted only in little
computational overhead.

@@

\includegraphics[width=.24\textwidth]{qualitative_twomoons_km}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_mean_nn}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_aggl}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_mst} \\ 
\caption{Comparison of $k$-Means (left), MeanNN (center left), single link
(center right) and ITM (right) on four synthetic datasets.  Without the need to
tune parameters, ITM can adjust to different cluster shapes.  MeanNN is able to
recover non-convex clusters (third row) but often produces similar results to
$k$-Means (second and last row). Single link clustering is very sensitive
to noise, as it does not take cluster size into account.}
\label{fig:qualitative}
\end{figure}

\section{Experiments}
We compared ITM to the popular $k$-Means algorithm 
\citep{macqueen1967some,lloyd1982least}, to the MeanNN algorithm of
\citet{faivishevsky2010nonparametric} and to single-link agglomerative
clustering~\citep{gower1969minimum}. The similarities between single-link
agglomerative clustering and the proposed MST-based optimization make it a 
good baseline for tree-based clustering approaches.

\enlargethispage{5mm}
A comparison of ITM, MeanNN and the baseline methods, $k$-Means and 
single link agglomerative clustering, in terms of their objective, 
optimization and complexity can be found in Table~\ref{nowotab}. 
We implemented the ITM clustering procedure as well as MeanNN in Python.  We
used the $k$-Means implementation available in the \textsc{scikit-learn}
library~\citep{pedregosa2011scikit}. The source
code is available online\footnote{\url{https://github.com/amueller/information-theoretic-mst}}.

\subsection{Experimental Setup}
For both $k$-Means and MeanNN, we restart the algorithm ten times using
different random initializations, keeping the result with the best objective
value. As ITM is deterministic there is no need for random restarts.  All of
the algorithms we compare work with a fixed number of clusters, which we set to
the number of classes in the dataset for all experiments.

@@

\begin{table}[t]
\centering
\begin{tabularx}{\linewidth}{@{\extracolsep{\fill}}lccc}
\toprule
Algorithm &     Objective &     Det.&      Complexity \\
\cmidrule{1-4}
$k$-Means &     $\displaystyle \sum_y \sum_{i, y_i = y} \| x_i - \mu_y \|^2$ & No  & $O(nk)$ per iteration%
%\footnote{The worst-time complexity of the Lloyd-Algorithm is exponential~\cite{arthur2006slow}, though online algorithm are observed to converge in linear time~\cite{bottou1995convergence}.}
\\
MeanNN &    $\displaystyle \sum_y \log\left(\frac{1}{|\mathbf{x}_y|}\sum_{i,j, y_i=y_j=y} \| x_i - x_j \|^2 \right)$ & No & $O(n^2)$ per iteration\\
Single Link &   -- &    Yes    &    $O(n\log n)$\\
ITM & $\displaystyle \sum_{y=0}^k d p(y) \log(\bar{L}_y  ) + p(y) \log{p(y)}$ & Yes & $O\bigl(\alpha(n) n \log n + nk\bigr)$\\
\bottomrule
\end{tabularx}
\caption{Comparing properties of related algorithms. Det.\ stands for Deterministic}\label{nowotab}
\end{table}

@@

\subsection{Qualitative Results}
\Figref{qualitative} shows qualitative results on three synthetic datasets.
For well separated, convex clusters, the four algorithms produce very similar
clusters (see top row).  If the structure of the data is more complex, the
advantage of the proposed method is apparent.  Note that there was no need to
specify any other parameters than the number of clusters to produce these
results.  It is also noteworthy that the results of MeanNN are very close to
those produces by $k$-Means in most cases. This similarity can be explained
by the close relation of the objective functions, listed in Table~\ref{nowotab}.

@@

We evaluated the experiments using the \emph{adjusted Rand index 
(ARI}~\citep{hubert1985comparing} a popular measure
of cluster quality~\citep{gomes2010discriminative,kamvar2003spectral}.
The Rand index~\citep{rand1971objective} between two clusterings counts on how
many pairs of points two clusterings agree. The adjusted Rand index contains a
calibration against chance performance.
We also measured \emph{normalized mutual information
(NMI)}~\citep{strehl2003cluster}, but do not report it here, as it resulted in
an identical ranking of the clustering algorithms.

@@

\begin{table}[t]
\centering
\begin{tabularx}{\linewidth}{@{\extracolsep{\fill}}lrrrccccc}
\toprule
\multicolumn{4}{c}{Dataset} &\multicolumn{4}{c}{Results}\\
\cmidrule{1-4}
\cmidrule{5-9}
Description &     n &    d &   k &  $k$-Means      & MeanNN        & SL            &   ITM        & ITM ID \\
\cmidrule{1-4}                   \cmidrule{5-5}    \cmidrule{6-6}  \cmidrule{7-7} \cmidrule{8-8} \cmidrule{9-9}
\textsc{digits}      &$ 1797$&$  64$&$ 10$& $    0.62 $&     $0.67$           &$0.10$         &$\mathbf{0.85}$       &$0.73$\\ 
\textsc{faces}       &$  400$&$4096$&$ 40$& $    0.41 $&     $0.49$           &$0.08$         &$0.02$           &$\B{0.54}$\\ 
\textsc{iris}        &$  150$&$   4$&$  3$& $    0.72 $&     $0.75$           &$0.55$         &$\B{0.88}$       &$0.88$\\
\textsc{usps}        &$ 9298$&$ 256$&$ 10$& $    0.52 $&     $0.54$           &$0.00$         &$   0.44$        &$\B{0.64}$\\ 
\textsc{vehicle}     &$  846$&$  18$&$  4$& $    0.10 $&     $0.09$           &$0.00$         &$\B{0.10}$       &$0.10$\\
\textsc{vowel}       &$  990$&$  10$&$ 11$& $    0.17 $&     $0.19$           &$0.00$         &$\B{0.20}$       &$0.19$\\
\textsc{waveform}    &$ 5000$&$  21$&$  2$& $\B{0.37} $&     $0.30$           &$0.00$         &$   0.23 $       &$0.23$\\
\textsc{MNIST}       &$70000$&$ 784$&$ 10$& $    0.37 $&     N/A$^\dagger$    &$0.00$         &$   0.50$        &$\B{0.77}$\\
\bottomrule
\end{tabularx}
\caption{Adjusted Rand Index of $k$-Means, MeanNN, single link agglomerative
    clustering and ITM on several datasets (higher is better). ITM ID refers to
    ITM using the estimated intrinsic dimensionality. The best score for each
    dataset is printed in bold.\\$^\dagger$\footnotesize We were unable to make
    MeanNN scale to 70000 data points, as storing the whole pairwise distance
    matrix seems necessary.\label{results}}
\end{table}

@@

\section{Summary}
In this chapter we proposed the use of a minimum spanning tree based,
non-parametric entropy estimator in information theoretic clustering, ITM\@.
Thereby we extended the work of \citet{faivishevsky2010nonparametric} to a more
flexible and efficient entropy estimate.  We proposed an approximate optimization
method by formulating the clustering problem as a search over graphs.
The resulting algorithm is deterministic and has sub-quadratic run time.
%
Empirical comparisons showed that the proposed method outperforms standard
algorithms and the non-parametric entropy based clustering of
\citep{faivishevsky2010nonparametric} on multiple benchmark datasets. We
demonstrated that ITM is able to detect non-convex clusters,
even in the presence of noise.
%
In contrast to other algorithms that can handle non-convex clusters, ITM has no
tuning parameters, as the objective presents a natural trade-off between
balancing cluster sizes and enforcing intra-cluster similarity.
%
A limitation of the proposed algorithm is that it is based on the assumption 
of an absolute continuous data distribution. We show that this limitation can
be overcome in practice by estimating the intrinsic dimensionality of the data. 
%
%In future work we plan to investigate a way to overcome this limitation, 
%for example by estimating the intrinsic dimensionality of the 
%data~\citep{pettis1979intrinsic}.
%
%We also investigate optimizations of the objective 
%\Eqref{graphobjective} that go beyond the proposed method. 
%Move-making algorithms seem a promising way to refine solutions 
%found by Algorithm~1. Branch and bound techniques could provide 
%an alternative approach.

>>>>

\section{Related Work}
The most commonly used clustering algorithm is the $k$-means algorithm, originally
investigated by \citet{macqueen1967some} and most commonly implemented using
Lloyd's algorithm \citep{macqueen1967some,lloyd1982least}. While
$k$-means often works well in practice, one of its main drawbacks is the
restriction in cluster shape. 
%
Clusters are given by the Voronoi tessellation of the cluster means and
therefore are always convex.  Another drawback is the non-determinism of the
procedure, caused by the dependence on random initialization.

Another widely used method is spectral clustering~\citep{shi2000normalized,ng2002spectral}, 
which solves a graph partitioning problem on a similarity
graph constructed from the data. While spectral clustering is much more
flexible than $k$-means it is quite sensitive to the particular choice of graph
construction and similarity measure. It is also computationally expensive to
compute, because clustering $n$ points requires computing the eigenvalues and
-vectors of an $n{\times}n$ matrix.

@@

Later, \citet{banerjee2005clustering} introduced the concept of Bregman
Information, generalizing mutual information of distributions, and showed how
this leads to a natural formulation of several clustering algorithms.
%
\citet{barber2006kernelized} constructed a soft clustering by using a parametric model of $p(Y\given X)$.
%
The framework of mutual information based clustering was extended to
non-parametric entropy estimates by \citet{faivishevsky2010nonparametric}.
They use a nearest neighbor based estimator of the mutual information, called
MeanNN, that takes into account all possible neighborhoods, therefore combining
global and local influences. The approximate mutual information is maximized
using local search over labels.
%discuss this later in detail
%While optimizing MeanNN allows for non-convex
%clusters, in practice the result is strongly biased towards compact, convex
%regions. This can be explained by a close connections of MeanNN to $k$-means,
%see Section~\ref{discussion} for an in-depth discussion.

Clustering algorithms based on minimum spanning trees have been studied early
on in the statistics community, due to their efficiency.  One of the earliest
methods is single-link agglomerative clustering~\citep{gower1969minimum}.
Single-link agglomerative clustering can be understood as a minimum spanning
tree-based approach in which the largest edge is removed until the desired
number of components is reached.  \citet{zahn1971graph} refined this criterion
by cutting edges that are longer than other edges in the vicinity. This
approach requires tuning several constants by hand. More recently,
\citet{grygorash2006minimum} proposed a hierarchical MST-based clustering
approach that iteratively cuts edges, merges points in the resulting
components, and rebuilds the spanning tree.
We limit our discussion to the most widely used algorithm from~\citep{gower1969minimum}.
%
\section[Clustering using Non-Parametric Entropy Estimates]{Clustering Using\linebreak Non-Parametric Entropy Estimates}
In general, the goal of clustering can be formulated as follows: 
given a finite collection of samples $\mathbf{x} = (x_1, \dotsc, x_n)$, we want to 
assign cluster-memberships $\mathbf{y} = (y_1, \dotsc, y_n ), y_i \in \{1, \dotsc k\}$ to these samples.
%
We adopt the viewpoint of information theoretic clustering of \citet{gokcay2002information}, 
where the $x_i$ are considered i.i.d{.} samples from a distribution $p(X)$, and the 
$y_i$ are found such that the mutual information $I(X, Y)$ between the distribution $p(X)$ and the
assigned labels $p(Y)$ is maximized.
We can rewrite this objective as 
\begin{align}
         I(X, Y)
        &= D_{\text{KL}}\bigl(p(X, Y) \parallel p(X) p(Y)\bigr) = H(X) - \sum_{y=1}^k p(Y{=}y) H(X \given Y {=} y)
\end{align}
where
\begin{itemize}
    \item $\displaystyle D_\text{KL}\bigl(p(X) \parallel q(X)\bigr)\, = \int_\mathcal{X} p(X) \ln\left(\frac{p(X)}{q(X)}\right)\,dX$ is the Kullback-Leibler divergence,
    \item $\displaystyle H(X)\, = \int_\mathcal{X} p(X) \ln\bigl(p(X)\bigr)\,dX$ is the differential entropy, and
    \item $\displaystyle H(X \given Y{=}y)\, = \int_\mathcal{X} p(X \given Y {=}y) \ln\bigl(p(X \given Y {=} y)\bigr)\,dX$ is the conditional differential entropy.
    \end{itemize}
Expressing the mutual information in terms of the entropy is convenient, since
the objective then decomposes over the values of $Y$. 
%
Additionally, $H(X)$ is independent of the distribution of $Y$ and therefore
does not influence the search over $\mathbf{y}$.

@@

We use Prim's algorithm combined with a ball tree data structure for distance
queries to compute the minimum spanning tree of the data.
While this procedure has no strong runtime guarantees, we found this faster
in practice than specialized algorithms for euclidean minimum spanning trees, 
which achieve a better theoretical runtime of $O\bigl(n \log(n) \alpha(n)\bigr)$ \citep{dtb2010}. Here
$\alpha$ is the inverse of the Ackerman function. The dynamic programming solution of
Algorithm~\ref{overall} has a run time of $O(n)$ per removed edge, leading to an overall
run time of $O\bigl(n \log(n) \alpha(n)+ nk\bigr)$. The $O(nk)$ term comes from a worst case
scenario, in which each step in the hierarchical clustering procedure only
splits off a constant number of points. In a more realistic setting, we expect
that the individual clusters are much smaller than the original dataset. In
this case, the $O(nk)$ term would improve to $O\bigl(n\log(k)\bigr)$.

@@

A particularly drastic example is the case of the dataset having less samples
than features. In this case, the data clearly lies even on a linear subspace
of the input space, and the dimensionality of the input space does not
accurately reflect the intrinsic dimensionality of the data.
We use the estimate analyzed by \citet{massoud2007manifold}.
In their method, for each data point $x$, a local estimate $\hat{d}(x)$ of the
dimensionality at $x$ is computed as
\begin{equation}
    \hat{d}(x) = \frac{\ln 2}{\ln\left(r_k(x) / r_{\lfloor k /2 \rfloor}(x)\right)}.
\end{equation}
Here $k$ is a fixed integer and $r_k(x)$ is the distance of $x$ from its $k$th neighbor.
We follow \citet{massoud2007manifold} and set $k =\lceil 2 \ln n\rceil$.
The final estimate $\hat{d}$ is then computed by averaging the estimates over all $x$
\begin{equation}
    \hat{d} = \frac{1}{n}\sum_{x \in X} \min\left(\hat{d}(x),\ d\right).
\end{equation}
We compute the density estimate once, prior to clustering, and then plug the
estimate $\hat{d}$ into \eqref{graphobjective} in place of $d$.
We found this estimate to work robustly and give sensible results for all
datasets we investigated. As we already used a ball tree data structure to
build the minimum spanning trees, we can reuse this structure to compute $r_k$.
Consequently, estimating the dimensionality resulted only in little
computational overhead.

@@

\includegraphics[width=.24\textwidth]{qualitative_twomoons_km}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_mean_nn}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_aggl}
\includegraphics[width=.24\textwidth]{qualitative_twomoons_mst} \\ 
\caption{Comparison of $k$-means (left), MeanNN (center left), single link
(center right) and ITM (right) on four synthetic datasets.  Without the need to
tune parameters, ITM can adjust to different cluster shapes.  MeanNN is able to
recover non-convex clusters (third row) but often produces similar results to
$k$-means (second and last row). Single link clustering is very sensitive
to noise, as it does not take cluster size into account.}
\label{fig:qualitative}
\end{figure}

\section{Experiments}
We compared ITM to the popular $k$-means algorithm 
\citep{macqueen1967some,lloyd1982least}, to the MeanNN algorithm of
\citet{faivishevsky2010nonparametric} and to single-link agglomerative
clustering~\citep{gower1969minimum}. The similarities between single-link
agglomerative clustering and the proposed MST-based optimization make it a 
good baseline for tree-based clustering approaches.

\enlargethispage{5mm}
A comparison of ITM, MeanNN and the baseline methods, $k$-means and 
single link agglomerative clustering, in terms of their objective, 
optimization and complexity can be found in Table~\ref{nowotab}. 
We implemented the ITM clustering procedure as well as MeanNN in Python.  We
used the $k$-means implementation available in the \textsc{scikit-learn}
library~\citep{pedregosa2011scikit}. The source
code is available online\footnote{\url{https://github.com/amueller/information-theoretic-mst}}.

\subsection{Experimental Setup}
For both $k$-means and MeanNN, we restart the algorithm ten times using
different random initializations, keeping the result with the best objective
value. As ITM is deterministic there is no need for random restarts.  All of
the algorithms we compare work with a fixed number of clusters, which we set to
the number of classes in the dataset for all experiments.

@@

\begin{table}[t]
\centering
\begin{tabularx}{\linewidth}{@{\extracolsep{\fill}}lccc}
\toprule
Algorithm &     Objective &     Det.&      Complexity \\
\cmidrule{1-4}
$k$-means &     $\displaystyle \sum_y \sum_{i, y_i = y} \| x_i - \mu_y \|^2$ & No  & $O(nk)$ per iteration%
%\footnote{The worst-time complexity of the Lloyd-Algorithm is exponential~\cite{arthur2006slow}, though online algorithm are observed to converge in linear time~\cite{bottou1995convergence}.}
\\
MeanNN &    $\displaystyle \sum_y \log\left(\frac{1}{|\mathbf{x}_y|}\sum_{i,j, y_i=y_j=y} \| x_i - x_j \|^2 \right)$ & No & $O(n^2)$ per iteration\\
Single Link &   -- &    Yes    &    $O(n\log n)$\\
ITM & $\displaystyle \sum_{y=0}^k d p(y) \log(\bar{L}_y  ) + p(y) \log{p(y)}$ & Yes & $O\bigl(\alpha(n) n \log n + nk\bigr)$\\
\bottomrule
\end{tabularx}
\caption{Comparing properties of related algorithms. Det.\ stands for Deterministic}\label{nowotab}
\end{table}

@@

\subsection{Qualitative Results}
\Figref{qualitative} shows qualitative results on three synthetic datasets.
For well separated, convex clusters, the four algorithms produce very similar
clusters (see top row).  If the structure of the data is more complex, the
advantage of the proposed method is apparent.  Note that there was no need to
specify any other parameters than the number of clusters to produce these
results.  It is also noteworthy that the results of MeanNN are very close to
those produces by $k$-means in most cases. This similarity can be explained
by the close relation of the objective functions, listed in Table~\ref{nowotab}.

@@

We evaluated the experiments using the \emph{adjusted Rand index 
(ARI}~\citep{hubert1985comparing} a popular measure
of cluster quality~\citep{gomes2010discriminative,kamvar2003spectral}.
The Rand index~\citep{rand1971objective} between two clusterings counts how
many pairs of points two clusterings agree. The adjusted Rand index contains a
calibration against chance performance.
We also measured \emph{normalized mutual information
(NMI)}~\citep{strehl2003cluster}, but do not report it here, as it resulted in
an identical ranking of the clustering algorithms.

@@

\begin{table}[t]
\centering
\begin{tabularx}{\linewidth}{@{\extracolsep{\fill}}lrrrccccc}
\toprule
\multicolumn{4}{c}{Dataset} &\multicolumn{4}{c}{Results}\\
\cmidrule{1-4}
\cmidrule{5-9}
Description &     n &    d &   k &  $k$-means      & MeanNN        & SL            &   ITM        & ITM ID \\
\cmidrule{1-4}                   \cmidrule{5-5}    \cmidrule{6-6}  \cmidrule{7-7} \cmidrule{8-8} \cmidrule{9-9}
\textsc{digits}      &$ 1797$&$  64$&$ 10$& $    0.62 $&     $0.67$           &$0.10$         &$\mathbf{0.85}$       &$0.73$\\ 
\textsc{faces}       &$  400$&$4096$&$ 40$& $    0.41 $&     $0.49$           &$0.08$         &$0.02$           &$\B{0.54}$\\ 
\textsc{iris}        &$  150$&$   4$&$  3$& $    0.72 $&     $0.75$           &$0.55$         &$\B{0.88}$       &$0.88$\\
\textsc{usps}        &$ 9298$&$ 256$&$ 10$& $    0.52 $&     $0.54$           &$0.00$         &$   0.44$        &$\B{0.64}$\\ 
\textsc{vehicle}     &$  846$&$  18$&$  4$& $    0.10 $&     $0.09$           &$0.00$         &$\B{0.10}$       &$0.10$\\
\textsc{vowel}       &$  990$&$  10$&$ 11$& $    0.17 $&     $0.19$           &$0.00$         &$\B{0.20}$       &$0.19$\\
\textsc{waveform}    &$ 5000$&$  21$&$  2$& $\B{0.37} $&     $0.30$           &$0.00$         &$   0.23 $       &$0.23$\\
\textsc{MNIST}       &$70000$&$ 784$&$ 10$& $    0.37 $&     N/A$^\dagger$    &$0.00$         &$   0.50$        &$\B{0.77}$\\
\bottomrule
\end{tabularx}
\caption{Adjusted Rand Index of $k$-means, MeanNN, single link agglomerative
    clustering and ITM on several datasets (higher is better). ITM ID refers to
    ITM using the estimated intrinsic dimensionality. The best score for each
    dataset is printed in bold.\\$^\dagger$\footnotesize We were unable to make
    MeanNN scale to 70000 data points, as storing the whole pairwise distance
    matrix seems necessary.\label{results}}
\end{table}

@@

\section{Summary}
In this chapter we proposed the use of a minimum spanning tree based,
non-parametric entropy estimator in information theoretic clustering, ITM\@.
Thereby we extended the work of \citet{faivishevsky2010nonparametric} to a more
flexible and efficient entropy estimate.  We proposed an approximate optimization
method by formulating the clustering problem as a search over graphs.
The resulting algorithm is deterministic and has sub-quadratic run time.
%
Empirical comparisons showed that the proposed method outperforms standard
algorithms and the non-parametric entropy based clustering of
\citet{faivishevsky2010nonparametric} on multiple benchmark datasets. We
demonstrated that ITM is able to detect non-convex clusters,
even in the presence of noise.
%
In contrast to other algorithms that can handle non-convex clusters, ITM has no
tuning parameters, as the objective presents a natural trade-off between
balancing cluster sizes and enforcing intra-cluster similarity.
%
A limitation of the proposed algorithm is that it is based on the assumption 
of an absolute continuous data distribution. We show that this limitation can
be overcome in practice by estimating the intrinsic dimensionality of the data. 
%
%In future work we plan to investigate a way to overcome this limitation, 
%for example by estimating the intrinsic dimensionality of the 
%data~\citep{pettis1979intrinsic}.
%
%We also investigate optimizations of the objective 
%\Eqref{graphobjective} that go beyond the proposed method. 
%Move-making algorithms seem a promising way to refine solutions 
%found by Algorithm~1. Branch and bound techniques could provide 
%an alternative approach.
